Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-22 paper.tex
Rough outlines of missing sections

@@

The VCF Zarr specification and scalable \texttt{vcf2zarr} conversion utility
provided here are a necessary starting point for such cloud-native
biobank repositories and open up many possibilities,
but significant investment and development would be needed 
to provide a viable alternative to standard bioinformatics workflows. 
Two initial directions for development, however, may quickly
yield sufficient results to both greatly improve researcher productivity on 
large, centrally managed datasets such as Genomics England
and motivate further research and development.
The first direction is to provide compatibility with existing 
workflows via a ``\texttt{vcztools}'' command line utility which 
implements a subset of \texttt{bcftools} functionality (such as 
\texttt{view} and \texttt{query}) on a VCF Zarr dataset.
Such a tool would speed up some common queries by orders 
of magnitude, and reduce the need for user orchestration of 
operations among manually split VCF chunks (large VCF datasets
are typically split into hundreds of files; see the Genomics
England case study). Datasets could then be hosted in cloud
object stores, while still presenting file-like semantics
for existing workflows. This could provide an evolutionary path,
allowing established analysis workflows to coexist with new Zarr-native
approaches, working from the same primary data.

The second natural direction for development is to create
these Zarr-native applications, which can take advantage
of the efficient data representation
across multiple programming languages (see Methods).
The Python data science ecosystem, in particular,
has a rich suite of powerful tools~\cite[e.g.][]{
mckinney2010data,
lam2015numba,
kluyver2016jupyter,
harris2020array,
virtanen2020scipy} 
and is increasingly
popular in recent biological applications~\citep[e.g.][]{
abdennur2020cooler,
rand2022bionumpy,
open2c2024bioframe,
hou2024admix}.
Xarray~\cite{hoyer2017xarray} provides a unified interface for working 
with multi-dimensional arrays
in Python, and libraries like Dask~\cite{rocklin2015dask}  
and Cubed~\cite{cubed2024} allow these operations to be
scaled out transparently across processors and clusters. 
This scaling is achieved by distributing calculations over 
grid-based array representations like Zarr, where chunks
provide the basic unit for parallel computation.
The VCF Zarr specification introduced here was created 
to facilitate work on a scalable genetics toolkit for 
Python~\cite{sgkit2024} built on Xarray. 
While the high-level facilities for distributed 
computation provided by Xarray 
are very powerful, they are not needed or indeed
appropriate in all contexts. Our benchmarks here 
illustrate that working at the lowest level, 
by sequentially applying optimised kernels on a chunk-by-chunk
basis is both straightforward to implement and highly performant.
Thus, a range of possibilities exist in which developers 
can build utilities using the VCF Zarr specification
using the appropriate level of abstraction and tool chain
on a case-by-case basis.

@@

@@

@@

@@

The VCF Zarr specification is available on GitHub at 
\url{https://github.com/sgkit-dev/vcf-zarr-spec/}.
All source code for running benchmarks, analyses and creating 
plots in this article is available at
\url{https://github.com/sgkit-dev/vcf-zarr-publication}.
Vcf2zarr is freely available under the terms of the Apache 2.0 
license as part of the \texttt{bio2zarr} 
suite (\url{https://github.com/sgkit-dev/bio2zarr/}) 
and can be installed from the Python Package Index
(\url{https://pypi.org/project/bio2zarr/}).

>>>>

\subsection{Interoperability: vcztools}
To provide compatibility with existing workflows we have 
developed the ``\texttt{vcztools}'' command line utility which 
implements a subset of \texttt{bcftools} functionality.
[TODO DETAILS]
% (such as 
% \texttt{view} and \texttt{query}) on a VCF Zarr dataset.
% Such a tool would speed up some common queries by orders 
% of magnitude, and reduce the need for user orchestration of 
% operations among manually split VCF chunks (large VCF datasets
% are typically split into hundreds of files; see the Genomics
% England case study). Datasets could then be hosted in cloud
% object stores, while still presenting file-like semantics
% for existing workflows. This could provide an evolutionary path,
% allowing established analysis workflows to coexist with new Zarr-native
% approaches, working from the same primary data.

\subsection{Interoperability: SAIGE proof-of-concept}
To demonstrate the potential for integrating VCF Zarr into existing utilities,
we developed a proof-of-concept extension of SAIGE~\cite{zhou2018efficiently,
zhou2020scalable}, a popular package for performing genome wide
association tests. Using the Tensorstore C++ library~\cite{tensorstore}
we added support for VCF Zarr via an additional backend class [some details 
about the implementation].

To test the performance of this implemenentation we compared the time required
to perform an association test using the simulated dataset of 10,000 samples 
from Fig~\ref{fig-data-storage}. We generated a phenotype using
\texttt{tstrait}~\citep{tagami2024tstrait}.

We compared the same analyis using the existing VCF, BCF and Savvy backends
to our prototype Zarr backend. 
[Times: BCF 3m, VCF6.5m, Savvy 39s, Zarr 55s]

% BCF:
% real 3m1.068s user 2m56.944s sys 0m2.941s

% VCF:
% real 6m30.335s user 6m25.193s sys 0m2.842s

% Savvy:
% real 0m39.660s user 0m37.498s sys 0m2.183s

% Zarr:
% real 0m55.957s user 1m10.375s sys 0m12.111s

@@

[TODO REWRITE]
The VCF Zarr specification and scalable \texttt{vcf2zarr} conversion utility
provided here are a necessary starting point for such cloud-native
biobank repositories and open up many possibilities,
but significant investment and development would be needed 
to provide a viable alternative to standard bioinformatics workflows. 
Two initial directions for development, however, may quickly
yield sufficient results to both greatly improve researcher productivity on 
large, centrally managed datasets such as Genomics England
and motivate further research and development.
The first direction is to provide compatibility with existing 
workflows via a ``\texttt{vcztools}'' command line utility which 
implements a subset of \texttt{bcftools} functionality (such as 
\texttt{view} and \texttt{query}) on a VCF Zarr dataset.
Such a tool would speed up some common queries by orders 
of magnitude, and reduce the need for user orchestration of 
operations among manually split VCF chunks (large VCF datasets
are typically split into hundreds of files; see the Genomics
England case study). Datasets could then be hosted in cloud
object stores, while still presenting file-like semantics
for existing workflows. This could provide an evolutionary path,
allowing established analysis workflows to coexist with new Zarr-native
approaches, working from the same primary data.

[TODO REWRITE]
The second natural direction for development is to create
these Zarr-native applications, which can take advantage
of the efficient data representation
across multiple programming languages (see Methods).
The Python data science ecosystem, in particular,
has a rich suite of powerful tools~\cite[e.g.][]{
mckinney2010data,
lam2015numba,
kluyver2016jupyter,
harris2020array,
virtanen2020scipy} 
and is increasingly
popular in recent biological applications~\citep[e.g.][]{
abdennur2020cooler,
rand2022bionumpy,
open2c2024bioframe,
hou2024admix}.
Xarray~\cite{hoyer2017xarray} provides a unified interface for working 
with multi-dimensional arrays
in Python, and libraries like Dask~\cite{rocklin2015dask}  
and Cubed~\cite{cubed2024} allow these operations to be
scaled out transparently across processors and clusters. 
This scaling is achieved by distributing calculations over 
grid-based array representations like Zarr, where chunks
provide the basic unit for parallel computation.
The VCF Zarr specification introduced here was created 
to facilitate work on a scalable genetics toolkit for 
Python~\cite{sgkit2024} built on Xarray. 
While the high-level facilities for distributed 
computation provided by Xarray 
are very powerful, they are not needed or indeed
appropriate in all contexts. Our benchmarks here 
illustrate that working at the lowest level, 
by sequentially applying optimised kernels on a chunk-by-chunk
basis is both straightforward to implement and highly performant.
Thus, a range of possibilities exist in which developers 
can build utilities using the VCF Zarr specification
using the appropriate level of abstraction and tool chain
on a case-by-case basis.

@@

\subsection{Local alleles}
[TODO Implementation of local alleles in vcf2zarr]

@@

\subsection{vcztools}
[TODO Implementation of vcztools]

@@

\subsection{AWS parallel Zarr prototype}
[TODO Implementation of prototype Zarr implementaion]

@@

The VCF Zarr specification is available on GitHub at 
\url{https://github.com/sgkit-dev/vcf-zarr-spec/}.
All source code for running benchmarks, analyses and creating 
plots in this article is available at
\url{https://github.com/sgkit-dev/vcf-zarr-publication}.
Vcf2zarr is freely available under the terms of the Apache 2.0 
license as part of the \texttt{bio2zarr} 
suite (\url{https://github.com/sgkit-dev/bio2zarr/}) 
and can be installed from the Python Package Index
(\url{https://pypi.org/project/bio2zarr/}).
Vcztools (\url{https://github.com/sgkit-dev/vcztools/}) 
is freely available under the terms of the Apache 2.0 
and can be installed from the Python Package Index
(\url{https://pypi.org/project/vcztools/}).
